Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-05-23 cordillera-cycle.tex
Remove the hi-res stuff for now.
Add preliminary list of authors.
Copy-paste methods from the previous paper.

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera have been more extensive than today. At the peak of glaciation, a
contiguous blanket of ice extended from Denali Peak in the north to the Puget
Sound in the south. This ice mass is known as the former Cordilleran ice sheet
(Fig.~\ref{fig:locmap}).

@@

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last
glacial cycle. However, former ice thickness, and in turn the ice sheet's
contribution to the last glacial maximum sea-level low-stand, remains poorly
constrained. Moreover, our understanding of the deglaciation phase is
restricted to a regional level, while little is known about the ice sheet
history preceding the last glacial maximum extent. Although time-evolving,
whole ice-sheet reconstructions of glacial advance and retreat patters are
available for the Laurentide and Eurasian ice sheet, such is not the case for
the Cordilleran ice sheet.

The present study aims to use a numerical ice sheet model calibrated against
field-based evidence to perform a quantitative reconstruction of the
Cordilleran ice sheet history through the last glacial cycle. Although
numerical modelling has been identified as a tool to improve our understanding
of Cordilleran ice sheet more than twenty years ago, the ubiquitously
mountainous topography of the region has presented a major challenge to its
application, which is only overcome by recent development of numerical ice
sheet models and underlying scientific tools. In addition to the topographic
complexity, strong climatic contrasts characteristic of the North American
Cordillera require the use of high-resolution temperature and precipitation
forcing as input data to the ice sheet model.

Here we use the Parallel Ice Sheet Model (PISM), forced by a high-resolution
climatology from the North American Regional Reanalysis (NARR), to model the
Cordilleran ice sheet evolution through the last glacial cycle. Because past climate
is mostly unknown, palaeoclimate forcing is simplistically applied as
six time-dependent temperature offset time-series. These time-series were
obtained by scaling six different palaeo-temperature reconstructions from
proxy records around the globe, including two \chem{\delta^{18}O} records from
Greenland ice cores, two \chem{\delta^{18}O} records from Antarctic ice cores,
and two alkenone unsaturation index records from Northwest Pacific oceanic
sediment cores. Model output is compared to geomorphological evidence in terms
of timing of glaciation, extent of glaciation and patterns of deglaciation.

@@

The simulations presented here were run using Parallel Ice Sheet Model (PISM,
development version~11b0a7f), an open-source, finite-difference, shallow ice
sheet model. The model takes basal topography, sea level, geothermal heat flux
and climate forcing, as input fields and computes the evolution of ice extent
and thickness in time, its thermal and dynamic state, and
the associated lithospheric response. Our modelling domain encompasses the
entire area covered by the Cordilleran ice sheet at the last glacial maximum.
It includes independent ice growing on the western Alaskan ranges and the
Brooks Range and intermediate expanses of ice-free terrain in northern Yukon
Territory and interior Alaska (Fig.~\ref{fig:locmap}).

To reconstruct the successive phases growth and decay of the last Cordilleran
ice sheet, palaeo-climatic conditions of the last glacial cycle are mimicked
by applying time-dependent temperature offsets derived from multiple
palaeo-temperature proxy records. Each simulation starts from assumed ice-free
conditions at -120\,kyr, and runs to present. These computations were
performed on 16 to 128 cores at the Swedish National Supercomputing
Center.

@@

@@

The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
SIA and SSA velocities are computed by finite difference methods on a~10\,km
resolution horizontal grid of 300 by 150 points (the modelling domain). Ice
softness depends on temperature and water content through an enthalpy
formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is
computed in three dimensions in up to 51 irregularly spaced layers within the
ice, and temperature is additionally computed in 31 regularly spaced layers in
bedrock to a~depth of 3\,km. Surface air temperature from the climate forcing
provides the upper boundary condition to the ice enthalpy model, and
a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary
condition to the bedrock thermal model. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial
variations in the geothermal heat flux may significantly affect basal
conditions \citep{pattyn-2010}, this study focuses on reconstructing the former
extent of the Cordilleran ice sheet, on which it has relatively little effect
\citep{rogozhina-etal-2012}.

\todo{Describe the new sliding law?}

A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is computed once at the beginning of the run,
as a~function of modern bed elevation, with lowest values occurring at low
elevations,

\begin{align}
  \phi =
  \begin{cases}
    15      & \text{for }       z<  0 \\
    z/15+15 & \text{for }    0 <z<200 \\
    45      & \text{for }  200 <z     \\
  \end{cases}
\end{align}

@@

Sea level is lowered as a function of time according to the SPECMAP
reconstruction, and basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{lingle-clark-1985,bueler-etal-2007}. It uses a lithosphere
density of 3300\,\unit{kg\,m^{-3}}, a flexural rigidity of $5 \times
10^{24}$\,\unit{N\,m} and a mantle viscosity of $1 \times
10^{21}$\,\unit{Pa\,s}. Due to the high mantle viscosity, there exists a time
lag between ice sheet growth and isostatic bedrock response.

@@

@@

The positive degree-day integral \citep{calov-greve-2005} is numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology and was
computed from daily temperature values from the North American Regional
Reanalysis \citep{data:narr} and excludes variability associated to the annual
temperature cycle itself \citet{seguinot-rogozhina-2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Atmospheric forcing of the model consists of a present-day monthly climatology
on the one hand, and time-dependent temperature offsets on the other hand. The
present-day climatology was computed from the near-surface air temperature and
precipitation rate fields of the North American Regional Reanalysis (NARR) over
the period xxxx--xxxx.

>>>>

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera have been more extensive than today. At the peak of glaciation, a
contiguous blanket of ice extended from Denali Peak in the north to the Puget
Sound in the south. This blanket is known as the former Cordilleran ice sheet.

@@

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last
glacial cycle. However, former ice thickness, and in turn the ice sheet's
contribution to the last glacial maximum sea-level low-stand, remains poorly
constrained. Moreover, our understanding of the deglaciation phase is
restricted to a regional level, while little is known about the ice sheet
history preceding the last glacial maximum extent. Although time-evolving,
whole ice-sheet reconstructions of glacial advance and retreat patters are
available for the Laurentide and Eurasian ice sheet, this is not the case for
the Cordilleran ice sheet.

The present study aims to use a numerical ice sheet model calibrated against
field-based evidence to attempt a quantitative reconstruction of the
Cordilleran ice sheet history through the last glacial cycle. Although
numerical modelling has been identified as a tool to improve our understanding
of Cordilleran ice sheet more than twenty years ago, the ubiquitously
mountainous topography of the region has presented a major challenge to its
application, which is only overcome by recent development of numerical ice
sheet models and underlying scientific tools. In addition to the topographic
complexity, strong climatic contrasts characteristic of the North American
Cordillera require the use of high-resolution temperature and precipitation
forcing as input data to the ice sheet model.

Here we use the Parallel Ice Sheet Model (PISM), forced by a high-resolution
climatology from the North American Regional Reanalysis, to model the
Cordilleran ice sheet through the last glacial cycle. Because past climate
is mostly unknown, palaeoclimate forcing is applied in the simplified form of
six time-dependent temperature offset time-series. These time-series were
obtained by scaling six different palaeo-temperature reconstructions from
proxy records around the globe, including two \chem{\delta^{18}O} records from
Greenland ice cores, two \chem{\delta^{18}O} records from Antarctic ice cores,
and two alkenone unsaturation index records from Northwest Pacific oceanic
sediment cores. The output from the model is compared to geomorphological
evidence in terms of timing of glaciation, extent of glaciation and patterns
of deglaciation.

@@

Given basal topography, sea level, geothermal heat flux and climate forcing,
the model computes ice extent and thickness, its thermal and dynamic state, and
the associated lithospheric response. Our modelling domain encompasses the
entire area covered by the Cordilleran ice sheet at LGM including independent
ice growing on the western Alaskan ranges and the Brooks Range and intermediate
expanses of ice-free terrain in northern Yukon Territory and interior Alaska
(Fig.~\ref{fig:locmap}).

As we aim to model glacial inception and growth of the Cordilleran ice sheet
towards a~configuration as last attained during the LGM, we mimic
palaeo-climatic conditions by applying constant temperature offsets
homogeneously over the modelling domain. Each simulation starts from ice-free
conditions and runs for 10\,kyr, a~time interval representative of the rapid
build-up of the last Cordilleran ice sheet from nearly ice-free to full glacial
conditions \citep{clague-1989,stroeven-etal-2010}. Our simulations are
performed in parallel on 32 cores at the Swedish National Supercomputing
Center.

@@

%SSA+SIA, pseudo-plastic sliding, temperature-dependent flow.
%Thermal model and deformation model.

@@

The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
SIA and SSA velocities are computed by finite difference methods on a~10\,km
resolution horizontal grid of 300 by 150 points (the modelling domain). Ice
softness depends on temperature and water content through an enthalpy
formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is
computed in three dimensions in up to 51 irregularly spaced layers within the
ice, and temperature is additionally computed in 11 regularly spaced layers in
bedrock to a~depth of 1\,km. Surface air temperature from the climate forcing
provides the upper boundary condition to the ice enthalpy model, and
a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary
condition to the bedrock thermal model. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial
variations in the geothermal heat flux may significantly affect basal
conditions \citep{pattyn-2010}, this study focuses on reconstructing the former
extent of the Cordilleran ice sheet, on which it has relatively little effect
\citep{rogozhina-etal-2012}.

A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 10 to 30{\degree}. It is computed once at the beginning of the run,
as a~function of modern bed elevation, with lowest values occurring at low
elevations,

\begin{align}
  \phi =
  \begin{cases}
    10      & \text{for }       z<  0 \\
    z/10+10 & \text{for }    0 <z<200 \\
    30      & \text{for }  200 <z     \\
  \end{cases}
\end{align}

@@

Sea level is lowered by 120\,m and basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{lingle-clark-1985,bueler-etal-2007}. It uses a lithosphere
density of 3300\,\unit{kg\,m^{-3}}, a flexural rigidity of $5 \times
10^{24}$\,\unit{N\,m} and a mantle viscosity of $1 \times
10^{21}$\,\unit{Pa\,s}. Due to the high mantle viscosity, there exists a time
lag between ice sheet growth and isostatic bedrock response.

@@

%Positive degree-day model including spatially and seasonally variable standard
%deviation (Fig.~\ref{fig:atm}).

@@

The positive degree-day integral \citep{calov-greve-2005} is numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is a~constant model parameter and was assigned
a~value of 3.07\,\unit{{\degree}C}, which corresponds to the mean summer (JJA),
model domain-averaged monthly standard deviation of daily mean temperature from
monthly mean temperature, as computed from North American Regional Reanalysis
data \citep{data:narr} in a~manner similar to \citet{seguinot-2013}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

%Input climatology from North American Regional Reanalysis according to
%sensitivity study from the previous paper (Fig.~\ref{fig:atm}).
